---
title: "Teaching"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
library(tidyverse)
library(kableExtra)

units <- read_csv("data/teaching.csv")
```

###  University Teaching 

Table 1 : List of Postgraduate Teachning at QUT - (CS60) Graduate Diploma in Medical Ultrasound, (CS61) Graduate Diploma in Cardiac Ultrasound

```{r , layout="l-body-outset"}
T1<- knitr::kable(units[1:5, ],
  booktabs = TRUE)

T1a<- column_spec(T1, 1, width = "4cm")
T1a
```

Table 2 : List of Undergradaute Teachning at QUT - (CS49) Bachelor of Radiation Therapy, (CS48) Bachelor of Medical Imaging (Honours)

```{r, layout="l-body-outset"}


T2<- knitr::kable(units[6:8, ],
  booktabs = TRUE)

T2a<- column_spec(T2, 1, width = "4cm")
T2a
```

### HDR - Honours Projects

Pressure pain thresholds in sub-elite athletes with and without plantar fasciopathy

Developmental dysplasia of the Hip - formulation of normal values for age using ultrasound

Dosimetric Impact of the Clarity Autoscan Ultrasound Probe on Perineal Dose in Prostate Cancer Patients 

### Invited Talks / Presentations

```{r}

```

