---
title: "Teaching"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
library(tidyverse)
library(kableExtra)

units <- read_csv("data/teaching.csv")
```

###  University Teaching 

List of postgraduate teaching at QUT - (CS60) Graduate Diploma in Medical Ultrasound, (CS61) Graduate Diploma in Cardiac Ultrasound

```{r , layout="l-body-outset"}
T1<- knitr::kable(units[1:5, ],
  booktabs = TRUE)

T1a<- column_spec(T1, 1, width = "4cm")
T1b<- column_spec(T1a, 2, width = "3cm")
T1b
```

List of undergraduate teaching at QUT - (CS49) Bachelor of Radiation Therapy, (CS48) Bachelor of Medical Imaging (Honours)

```{r, layout="l-body-outset"}


T2<- knitr::kable(units[6:8, ],
  booktabs = TRUE)

T2a<- column_spec(T2, 1, width = "4cm")
T2b<- column_spec(T2a, 2, width = "3cm")
T2b
```

### HDR & Honours Projects

Pressure pain thresholds in sub-elite athletes with and without plantar fasciopathy

Developmental dysplasia of the Hip - formulation of normal values for age using ultrasound

Dosimetric Impact of the Clarity Autoscan Ultrasound Probe on Perineal Dose in Prostate Cancer Patients 

### Invited Talks / Presentations

*Longitudinal analysis of placental stiffness using ultrasound - Towards the Digital Placenta*. [QUT CBT Conference](https://research.qut.edu.au/cbt/conference/), Brisbane, 11-12 November 2021

*Modelling the biomechanical properties of the placenta using sonoelastography*. Placental Consortium Seminar [Stillbirth Centre of Research Excellence](https://stillbirthcre.org.au/), Brisbane, 23 March November 2022 [slides](./data/Placental_oral.pdf)

*Understanding the reliability and consistency of Ultrasound dervied elastrography measures of the human placenta*. PhD Final Seminar [Queensland University of Technology (QUT)](https://www.qut.edu.au/), Brisbane, 15 April 2022 [slides](./data/PHD Thesis Final Seminar.pdf)

*Using non-invasive ultrasound elastography to monitor structural changes of the placenta*. QLD Family Cohort Symposium [QLD Family Cohort / Mater Research](https://www.qldfamilycohort.org/), Translational Research Institue, Brisbane, 20 May 2022 

```{r}

```

